\tit

### 2021-02-17 paper.tex
box-dot

\ifacm\else
%% Already in acmart
\RequirePackage{amssymb, unicode-math}
\fi
\RequirePackage{newunicodechar, stmaryrd, setspace, comment}

@@

%% I prefer ∅ in XITSMath-Regular, but I don't know how to make setmathfont
%% work with acmart.
\ifacm
\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
\else
\setmathfont{XITSMath-Regular.otf}
\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
\fi

@@

A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages.
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

\sectionl{Predicate Algebra}

@@

Fortunately, there is a simple way to encapsulate indexed types into simple (non-indexed) types, namely existential quantification.
For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
For a type family {\AB F \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
For instance, consider language union from \figref{symbolic-api}:
\begin{code}
% Copied from latex/Symbolic.tex
\>[2]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∪\AgdaUnderscore{}}}%
\>[7]\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Lang}%
\>[15]\AgdaGeneralizable{P}%
\>[18]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaGeneralizable{Q}%
\>[28]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaGeneralizable{P}%
\>[39]\AgdaOperator{\AgdaFunction{◬.∪}}%
\>[44]\AgdaGeneralizable{Q}\AgdaSymbol{)}\<%
\end{code}
Referring to the type and operations of \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'', define \agda{wrapped-Lang}
Then wrap union as follows:
\agda{wrapped-union}
Note that the type-level language components of the arguments and result are left to be inferred by the compiler.
\rnc\source{Existential}
Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, this time with the inferable components filled in for clarity.
Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.

>>>>

\ifacm\else
%% Already in acmart
\RequirePackage{amssymb, unicode-math}
\fi
\RequirePackage{newunicodechar, stmaryrd, setspace, comment, scalerel}

@@

%% I prefer ∅ in XITSMath-Regular, but I don't know how to make setmathfont
%% work with acmart.
\ifacm
\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
\newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
\else
\setmathfont{XITSMath-Regular.otf}
\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
\newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
\fi

@@

A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

@@

\subsectionl{Predicate Algebra}

@@

Fortunately, there is a simple way to encapsulate indexed types into simple (non-indexed) types, namely existential quantification.
For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
For a type family {\AB F \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
For instance, consider language union from \figref{symbolic-api}:
\begin{code}
% Copied from latex/Symbolic.tex
\>[2]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∪\AgdaUnderscore{}}}%
\>[7]\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Lang}%
\>[15]\AgdaGeneralizable{P}%
\>[18]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaGeneralizable{Q}%
\>[28]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaGeneralizable{P}%
\>[39]\AgdaOperator{\AgdaFunction{◬.∪}}%
\>[44]\AgdaGeneralizable{Q}\AgdaSymbol{)}\<%
\end{code}
Referring to the type and operations of \secref{Specifying Languages} and \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
Then wrap union as follows:
\agda{wrapped-union}
The type-level language components of the arguments and result are given explicitly here but could instead be inferred by the compiler.
\rnc\source{Existential}
Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with inferable components given explicitly for clarity.
Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.
